---
title: "Milestone 1"
author: "Yesh Onipede"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setting up data paths 
```{r}
merged_data <- read.csv("data/Gin_LiquorSales_Education.csv")

```

Loading packages 
```{r}
library(forecast)
library(ggplot2)
library(tidyr)
library(dplyr)
library(cowplot)


```
Creating an aggregated table and plotting 
```{r}
# Convert the date column to Date format 
merged_data$date <- as.Date(merged_data$date)

# Group the data by 'date and summarize to calculate the sum of each variable for each date
aggregated_data <- merged_data %>%
  group_by(date) %>%
  summarize(total_sale_dollars = sum(sale.dollars),
            total_sale_volume = sum(sale.volume),
            total_sale_bottles = sum(sale.bottles))

# Print the aggregated data frame
print(aggregated_data)

# Plotting
ggplot(aggregated_data, aes(x = date)) +
  geom_line(aes(y = total_sale_dollars, color = "Sales ($)")) +
  geom_line(aes(y = total_sale_volume, color = "Volume (Liters)")) +
  geom_line(aes(y = total_sale_bottles, color = "Bottles Sold")) +
  labs(x = "Date", y = "Value", color = "Legend") +
  ggtitle("Aggregated Sales, Volume, and Bottles Sold Over Time") +
  theme_minimal()


```

```{r}


# Extract Year from the date column
merged_data$year <- (lubridate::year(merged_data$date))

# Creating an aggregated dataset based on year and quarter
aggregated_data <- merged_data %>%
  group_by(year, quarter) %>%
  summarize(total_sale_dollars = sum(sale.dollars),
            total_sale_volume = sum(sale.volume),
            total_sale_bottles = sum(sale.bottles)) %>%
  mutate(year_quarter = paste0(as.character(year), quarter))

# Print the updated aggregated data frame
print(aggregated_data)






```

Creating time series objects
```{r}

# Create a date column using the year and quarter components
aggregated_data$date <- as.Date(paste0(aggregated_data$year, "-", aggregated_data$quarter), format = "%Y-%m")

# Convert to quarterly time series for sales (dollars)
ts_data_dollars <- ts(aggregated_data$total_sale_dollars, frequency = 4, start = c(min(aggregated_data$year), 1))

# Convert to quarterly time series for sales (bottles)
ts_data_bottles <- ts(aggregated_data$total_sale_bottles, frequency = 4, start = c(min(aggregated_data$year), 1))

# Convert to quarterly time series for sales (volume)
ts_data_volume <- ts(aggregated_data$total_sale_volume, frequency = 4, start = c(min(aggregated_data$year), 1))


```



Forecasting with an snaive model (dollars)
```{r}
# Fit the snaive model
snaive_model_dollars <- snaive(ts_data_dollars)

# Forecast for the next 4 quarters
forecast_result_dollars<- forecast(snaive_model_dollars, h = 8)

# Print the forecast
print(forecast_result_dollars)

autoplot(forecast_result_dollars) + 
  autolayer(ts_data_dollars, series = "Seasonal naive model") +
  xlab("Year") +
  ylab("Total Sale Dollars") +
  ggtitle("Forecast of Total Sale Dollars") +
  guides(colour = guide_legend(title = "Forecast"))  

```
#ARIMA forecast but I am not sure why this is not working yet since I am getting the same point forecast for every upcoming quarter 
```{r}
# Fit the ARIMA model
arima_model <- auto.arima(ts_data_dollars)

# Forecast for the next 8 quarters
forecast_result_arima <- forecast(arima_model, h = 8)

# Print the ARIMA forecast
print(forecast_result_arima)

# Plot the ARIMA forecast
autoplot(forecast_result_arima) + 
  autolayer(ts_data_dollars, series = "Actual") +
  xlab("Year") +
  ylab("Total Sale Dollars") +
  ggtitle("ARIMA Forecast of Total Sale Dollars") +
  guides(colour = guide_legend(title = "Forecast"))

```

Forecasting with an snaive model (bottles)
```{r}
# Fit the snaive model
snaive_model_bottles <- snaive(ts_data_bottles)

# Forecast for the next 4 quarters
forecast_result_bottles <- forecast(snaive_model_bottles, h = 8)

# Print the forecast
print(forecast_result_bottles)

autoplot(forecast_result_bottles) + 
  autolayer(ts_data_bottles, series = "Seasonal Naive Model") +
  xlab("Year") +
  ylab("Total Sale (Bottles)") +
  ggtitle("Forecast of Total Sale (Bottles)") +
  guides(colour = guide_legend(title = "Forecast"))  
```

Forecasting with an snaive (volume)
```{r}
# Fit the snaive model
snaive_model_volume <- snaive(ts_data_volume)

# Forecast for the next 4 quarters
forecast_result_volume <- forecast(snaive_model_volume, h = 8)

# Print the forecast
print(forecast_result_volume)

autoplot(forecast_result_volume) + 
  autolayer(ts_data_volume, series = "Seasonal Naive Model") +
  xlab("Year") +
  ylab("Total Sale (Volume)") +
  ggtitle("Forecast of Total Sale (Volume)") +
  guides(colour = guide_legend(title = "Forecast"))  

```

Create an aggregated table by county
```{r}
# Extract Year from the date column
merged_data$year <- lubridate::year(merged_data$date)

# Creating an aggregated dataset based on year, quarter, and county
aggregated_data_county <- merged_data %>%
  group_by(year, quarter, county) %>%
  summarize(total_sale_dollars = sum(sale.dollars),
            total_sale_volume = sum(sale.volume),
            total_sale_bottles = sum(sale.bottles)) %>%
  mutate(year_quarter = paste0(year, quarter))

# Print the updated aggregated data frame
print(aggregated_data_county)

```

i.e. Forecasating for a specific county 
```{r}

# Filter the aggregated data to include only rows where the county is "Adair"
aggregated_data_adair <- filter(aggregated_data, county == "Adair")

# Create a date column using the year and quarter components
aggregated_data_adair$date <- as.Date(paste0(aggregated_data_adair$year, "-", aggregated_data_adair$quarter), format = "%Y-%m")

# Convert to quarterly time series for sales (dollars)
ts_data_dollars_adair <- ts(aggregated_data_adair$total_sale_dollars, frequency = 4, start = c(min(aggregated_data_adair$year), 1))

# Convert to quarterly time series for sales (bottles)
ts_data_bottles_adair <- ts(aggregated_data_adair$total_sale_bottles, frequency = 4, start = c(min(aggregated_data_adair$year), 1))

# Convert to quarterly time series for sales (volume)
ts_data_volume_adair <- ts(aggregated_data_adair$total_sale_volume, frequency = 4, start = c(min(aggregated_data_adair$year), 1))

```

Forecasating for dollar sales for adair
```{r}

forecast_dollars_adair <- snaive(ts_data_dollars_adair)

# Print the forecasted values
print(forecast_dollars_adair)

# Plot the forecasted values
autoplot(forecast_dollars_adair) +
  ggtitle("Forecasted Dollar Sales for County Adair (Seasonal Naive Method)") +
  xlab("Year") +
  ylab("Dollar Sales")
```