---
title: "Milestone 1"
author: "Yesh Onipede"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setting up data paths 
```{r}
#setwd('/Users/yeshimonipede/Desktop/BC_Spring2024')
merged_data <- read.csv("/Users/yeshimonipede/Desktop/BC_Spring2024/Iowa_LiquorSales_Education/Gin_LiquorSales_Education.csv")
```

Loading packages 
```{r}
library(forecast)
library(ggplot2)
library(tidyr)
library(dplyr)
library(cowplot)


```
Creating an aggregated table and plotting 
```{r}
# Convert the date column to Date format 
merged_data$date <- as.Date(merged_data$date)

# Group the data by 'date and summarize to calculate the sum of each variable for each date
aggregated_data <- merged_data %>%
  group_by(date) %>%
  summarize(total_sale_dollars = sum(sale.dollars),
            total_sale_volume = sum(sale.volume),
            total_sale_bottles = sum(sale.bottles))

# Print the aggregated data frame
print(aggregated_data)

# Plotting
ggplot(aggregated_data, aes(x = date)) +
  geom_line(aes(y = total_sale_dollars, color = "Sales ($)")) +
  geom_line(aes(y = total_sale_volume, color = "Volume (Liters)")) +
  geom_line(aes(y = total_sale_bottles, color = "Bottles Sold")) +
  labs(x = "Date", y = "Value", color = "Legend") +
  ggtitle("Aggregated Sales, Volume, and Bottles Sold Over Time") +
  theme_minimal()


```

```{r}


# Extract Year from the date column
merged_data$year <- (lubridate::year(merged_data$date))

# Creating an aggregated dataset based on year and quarter
aggregated_data <- merged_data %>%
  group_by(year, month) %>%
  summarize(total_sale_dollars = sum(sale.dollars),
            total_sale_volume = sum(sale.volume),
            total_sale_bottles = sum(sale.bottles)) 
  #mutate(year_quarter = paste0(as.character(year), quarter))

# Print the updated aggregated data frame
print(aggregated_data)






```

Creating time series objects
```{r}

# Create a date column using the year and quarter components
aggregated_data$date <- as.Date(paste0(aggregated_data$year, "-", aggregated_data$month), format = "%Y-%m")

# Convert to quarterly time series for sales (dollars)
ts_data_dollars <- ts(aggregated_data$total_sale_dollars, frequency = 12, start = c(min(aggregated_data$year), 1))

# Convert to quarterly time series for sales (bottles)
ts_data_bottles <- ts(aggregated_data$total_sale_bottles, frequency = 12, start = c(min(aggregated_data$year), 1))

# Convert to quarterly time series for sales (volume)
ts_data_volume <- ts(aggregated_data$total_sale_volume, frequency = 12, start = c(min(aggregated_data$year), 1))


```



Forecasting with an snaive model (dollars)
```{r}
# Fit the snaive model
snaive_model_dollars <- snaive(ts_data_dollars)

# Forecast for the next 4 quarters
forecast_result_dollars<- forecast(snaive_model_dollars, h = 8)

# Print the forecast
print(forecast_result_dollars)

autoplot(forecast_result_dollars) + 
  autolayer(ts_data_dollars, series = "Seasonal naive model") +
  xlab("Year") +
  ylab("Total Sale Dollars") +
  ggtitle("Forecast of Total Sale Dollars") +
  guides(colour = guide_legend(title = "Forecast"))  

```
#ARIMA forecast but I am not sure why this is not working yet since I am getting the same point forecast for every upcoming quarter 
```{r}
# Fit the ARIMA model
arima_model <- auto.arima(ts_data_dollars)

# Forecast for the next 8 quarters
forecast_result_arima <- forecast(arima_model, h = 24)

# Print the ARIMA forecast
print(forecast_result_arima)

# Plot the ARIMA forecast
autoplot(forecast_result_arima) + 
  autolayer(ts_data_dollars, series = "Actual") +
  xlab("Year") +
  ylab("Total Sale Dollars") +
  ggtitle("ARIMA Forecast of Total Sale Dollars") +
  guides(colour = guide_legend(title = "Forecast"))

```
```{r}
# Fit SES model
ses_model <- ets(ts_data_dollars)

# Forecast for the next 8 quarters
forecast_result_ses <- forecast(ses_model, h = 24)

# Print the SES forecast
print(forecast_result_ses)

# Plot the SES forecast
autoplot(forecast_result_ses) + 
  autolayer(ts_data_dollars, series = "Actual") +
  xlab("Year") +
  ylab("Total Sale Dollars") +
  ggtitle("SES Forecast of Total Sale Dollars") +
  guides(colour = guide_legend(title = "Forecast"))

```



Forecasting with an snaive model (bottles)
```{r}
# Fit the snaive model
snaive_model_bottles <- snaive(ts_data_bottles)

# Forecast for the next 4 quarters
forecast_result_bottles <- forecast(snaive_model_bottles, h = 8)

# Print the forecast
print(forecast_result_bottles)

autoplot(forecast_result_bottles) + 
  autolayer(ts_data_bottles, series = "Seasonal Naive Model") +
  xlab("Year") +
  ylab("Total Sale (Bottles)") +
  ggtitle("Forecast of Total Sale (Bottles)") +
  guides(colour = guide_legend(title = "Forecast"))  
```

Forecasting with an snaive (volume)
```{r}
# Fit the snaive model
snaive_model_volume <- snaive(ts_data_volume)

# Forecast for the next 4 quarters
forecast_result_volume <- forecast(snaive_model_volume, h = 8)

# Print the forecast
print(forecast_result_volume)

autoplot(forecast_result_volume) + 
  autolayer(ts_data_volume, series = "Seasonal Naive Model") +
  xlab("Year") +
  ylab("Total Sale (Volume)") +
  ggtitle("Forecast of Total Sale (Volume)") +
  guides(colour = guide_legend(title = "Forecast"))  

```

Create an aggregated table by county
```{r}
# Extract Year from the date column
merged_data$year <- lubridate::year(merged_data$date)

# Creating an aggregated dataset based on year, quarter, and county
aggregated_data_county <- merged_data %>%
  group_by(year, month, county) %>%
  summarize(total_sale_dollars = sum(sale.dollars),
            total_sale_volume = sum(sale.volume),
            total_sale_bottles = sum(sale.bottles)) %>%
  mutate(year_quarter = paste0(year, month))

# Print the updated aggregated data frame
print(aggregated_data_county)

```

i.e. Forecasating for a specific county 
```{r}

# Filter the aggregated data to include only rows where the county is "Adair"
aggregated_data_adair <- filter(aggregated_data_county, county == "Adair")

# Create a date column using the year and quarter components
aggregated_data_adair$date <- as.Date(paste0(aggregated_data_adair$year, "-", aggregated_data_adair$month), format = "%Y-%m")

# Convert to quarterly time series for sales (dollars)
ts_data_dollars_adair <- ts(aggregated_data_adair$total_sale_dollars, frequency = 12, start = c(min(aggregated_data_adair$year), 1))

# Convert to quarterly time series for sales (bottles)
ts_data_bottles_adair <- ts(aggregated_data_adair$total_sale_bottles, frequency = 12, start = c(min(aggregated_data_adair$year), 1))

# Convert to quarterly time series for sales (volume)
ts_data_volume_adair <- ts(aggregated_data_adair$total_sale_volume, frequency = 12, start = c(min(aggregated_data_adair$year), 1))

```

Forecasating for dollar sales for adair
```{r}

arima_dollars_adair <- auto.arima(ts_data_dollars_adair)

forecast_adair<-forecast(arima_dollars_adair,h=12)
# Print the forecasted values
print(forecast_adair)

# Plot the forecasted values
autoplot(forecast_adair) +
  ggtitle("Forecasted Dollar Sales for County Adair (Seasonal Naive Method)") +
  xlab("Year") +
  ylab("Dollar Sales")
```


i.e. Forecasating for a specific county 
```{r}

# Filter the aggregated data to include only rows where the county is "Adair"
aggregated_data_adams <- filter(aggregated_data_county, county == "Adams")

# Create a date column using the year and quarter components
aggregated_data_adams$date <- as.Date(paste0(aggregated_data_adams$year, "-", aggregated_data_adams$month), format = "%Y-%m")

# Convert to quarterly time series for sales (dollars)
ts_data_dollars_adams<- ts(aggregated_data_adams$total_sale_dollars, frequency = 12, start = c(min(aggregated_data_adams$year), 1))

# Convert to quarterly time series for sales (bottles)
ts_data_bottles_adams <- ts(aggregated_data_adams$total_sale_bottles, frequency = 12, start = c(min(aggregated_data_adams$year), 1))

# Convert to quarterly time series for sales (volume)
ts_data_volume_adams <- ts(aggregated_data_adams$total_sale_volume, frequency = 12, start = c(min(aggregated_data_adams$year), 1))

```

Forecasating for dollar sales for adair
```{r}

arima_dollars_adams <- auto.arima(ts_data_dollars_adams)

forecast_adams<-forecast(arima_dollars_adams,h=12)
# Print the forecasted values
print(forecast_adams)

# Plot the forecasted values
autoplot(forecast_adams) +
  ggtitle("Forecasted Dollar Sales for County Adair (Seasonal Naive Method)") +
  xlab("Year") +
  ylab("Dollar Sales")
```


Looping through each county and creating a time series of sales in terms of bottles and in terms of volume
```{r}
create_time_series <- function(data, target_county) {
  # Filter the data for the specific county
  county_data <- filter(data, county == target_county)


  # Convert to monthly time series for sales (dollars)
    ts_sales <- ts(county_data$total_sale_dollars, frequency = 12, start = c(2016,1), end = c(2017,12))
  
  # Convert to monthly time series for volume
    ts_volume <- ts(county_data$total_sale_volume, frequency = 12, start = c(2016,1), end = c(2017,12))
  
  # Return a list containing the time series objects
  return(list(
    ts_sales = ts_sales,
    ts_volume = ts_volume
  ))
}

# Get unique county names
county_names <- unique(aggregated_data_county$county)

# Create an empty list to store time series data
time_series_list <- list()

# Loop through each county and create time series
for (county in county_names) {
  time_series_list[[county]] <- create_time_series(aggregated_data_county, county)
}


```
ARIMA model across all counties and outputting the counties that are predicted to have the largest increase over the next two years 
```{r}
library(forecast)

# Function to fit ARIMA model and forecast liquor sales
fit_arima_forecast <- function(ts_sales) {
  # Fit ARIMA model
  arima_model <- auto.arima(ts_sales)
  
  # Forecast liquor sales for the next two years
  forecast_values <- forecast(arima_model, h = 24)  # 24 months = 2 years
  
  # Return the forecasted values
  return(forecast_values)
}

# Initialize a list to store forecasted sales for each county
forecasted_sales <- list()

# Loop through each county
for (county in county_names) {
  # Filter data for the current county
  county_data <- time_series_list[[county]]
  
  # Fit ARIMA model and forecast liquor sales
  forecast_values <- fit_arima_forecast(county_data$ts_sales)
  
  # Store the forecasted values for the county
  forecasted_sales[[county]] <- forecast_values
}

# Calculate the increase in liquor sales for each county
increase_sales <- sapply(forecasted_sales, function(forecast_values) {
  # Extract the forecasted values for the next two years
  forecast_sales <- forecast_values$mean
  
  # Calculate the increase in liquor sales
  increase <- forecast_sales[length(forecast_sales)] - forecast_sales[1]
  
  return(increase)
})

# Find the counties with the largest forecasted increase in liquor sales
top_counties <- names(increase_sales)[order(increase_sales, decreasing = TRUE)][1:5]

# Print the top counties
print(top_counties)

# Autoplot the forecasts for the top counties
for (county in top_counties) {
  # Plot the forecast
  autoplot(forecasted_sales[[county]], main = paste("Forecast for", county), ylab = "Sales")
}

```
```{r}
library(forecast)

# Function to fit seasonal naïve (snaive) model and forecast liquor sales
fit_snaive_forecast <- function(ts_sales) {
  # Fit seasonal naïve (snaive) model
  snaive_model <- snaive(ts_sales)
  
  # Forecast liquor sales for the next two years
  forecast_values <- forecast(snaive_model, h = 24)  # 24 months = 2 years
  
  # Return the forecasted values
  return(forecast_values)
}

# Initialize a list to store forecasted sales for each county using snaive model
forecasted_sales_snaive <- list()

# Loop through each county
for (county in county_names) {
  # Filter data for the current county
  county_data <- time_series_list[[county]]
  
  # Fit snaive model and forecast liquor sales
  forecast_values <- fit_snaive_forecast(county_data$ts_sales)
  
  # Store the forecasted values for the county
  forecasted_sales_snaive[[county]] <- forecast_values
}

# Calculate the increase in liquor sales for each county using snaive forecasts
increase_sales_snaive <- sapply(forecasted_sales_snaive, function(forecast_values) {
  # Extract the forecasted values for the next two years
  forecast_sales <- forecast_values$mean
  
  # Calculate the increase in liquor sales
  increase <- forecast_sales[length(forecast_sales)] - forecast_sales[1]
  
  return(increase)
})

# Find the counties with the largest forecasted increase in liquor sales using snaive forecasts
top_counties_snaive <- names(increase_sales_snaive)[order(increase_sales_snaive, decreasing = TRUE)][1:5]

# Print the top counties based on snaive forecasts
print(top_counties_snaive)

# Autoplot the forecasts for the top counties using snaive forecasts
for (county in top_counties_snaive) {
  # Plot the forecast
  autoplot(forecasted_sales_snaive[[county]], main = paste("Seasonal Naïve Forecast for", county), ylab = "Sales")
}

```
